---
title: "Project 2"
format: html
editor: visual
---

This markdown document will include a description of the dataset, detailed data combining/cleaning activities, and exploratory data analysis with summary tables and plots. These steps are prerequisite to creating the associated Shiny dashboard app.

```{r}
#| message: false
library(tidyr)
library(dplyr)
library(readr)
```


## 1. Description of Dataset

The dataset for this analysis comes from the Tennis Match Charting Project (MCP), available at <https://github.com/JeffSackmann/tennis_MatchChartingProject.git>. The MCP aims to build a detailed public record of professional tennis matches. MCP data covers over 11,000 singles matches including both match-summary statistics and shot-by-shot detail. The project is open to the public both for accessing the data and for contributing new charted matches. This analysis uses only the summary data.

Four raw data .csv files are downloaded from the GitHub repo and placed in `RawData/`:

- `charting-m-matches.csv`

- `charting-m-stats-Overview.csv`

- `charting-w-matches.csv`

- `charting-w-stats-Overview.csv`

"m" and "w" indicate men's and women's professional tennis tour matches, respectively. "...-matches.csv" files contain general information on the matches played such as players, date, tournament, and surface. "...-stats-Overview.csv" files contain more detailed match performance information such as points won, aces, unforced errors, etc. Each observation across all four files contains a match_id variable, indicating the specific match observed. This variable will be used to join the data.

## 2. Data Combination and Cleaning

Combining and cleaning the data will involve the following general steps:

1.  Combine all four raw data files into a single tibble, adding a categorical variable to indicate 'M' or 'W' tour.

2.  Delete variables and/or observations that will not be used in subsequent analysis

3.  Confirm correct format/type for all remaining variables

4.  Add new variables calculated from existing variables

The variables for the "-matches.csv" files are as follows:

| Variable | Type | Notes |
|------------------|------------------|-------------------------------------|
| match_id | Identifier | Will use for join with overview files |
| Player 1 | Categorical | Name of Player 1 |
| Player 2 | Categorical | Name of Player 2 |
| Pl 1 hand | Categorical | Handed for Player 1 |
| Pl 2 hand | Categorical | Handed for Player 2 |
| Date | Numerical | Currently stored as YYYYMMDD integer. |
| Tournament | Categorical | Name of Tournament |
| Round | Categorical | Round of the Tournament (i.e. F, SF, R64, etc.) |
| Time | Categorical | Many missing values and inconsistent formatting. Delete later. |
| Court | Categorical | Name/number of court. Many missing values. Delete later. |
| Surface | Categorical | Hard, Clay or Grass |
| Umpire | Categorical | Name of umpire. Many missing values. Delete later. |
| Best of | Numerical | Best of 3 or Best of 5. |
| Final TB? | Categorical | Meaning unclear. Delete later. |
| Charted by | Categorical | Name of data contributer. Delete later. |

The variables for the "-Overview.csv" files are as follows:

| Variable | Type | Notes |
|-----------------|-----------------|--------------------------------------|
| match_id | Identifier | Will use for join with overview files |
| player | Categorical | Name of player |
| set | Categorical | Stats by set number. Filter later to keep only "Total" |
| serve_pts | Numerical | Points served |
| aces | Numerical | Aces |
| dfs | Numerical | Double faults |
| first_in | Numerical | First serves in play |
| first_won | Numerical | First serve points won |
| second_in | Numerical | Second serves in play |
| second_won | Numerical | Second serve points won |
| bk_pts | Numerical | Break points won |
| bp_saved | Numerical | Break points saved |
| return_pts | Numerical | Points returned |
| return_pts_won | Numerical | Return points won |
| winners | Numerical | Winners |
| winners_fh | Numerical | Forehand winners |
| winners_bh | Numerical | Backhand winners |
| unforced | Numerical | Unforced errors |
| unforced_fh | Numerical | Forehand unforced errors |
| unforced_bh | Numerical | Backhand unforced errors |

Now to begin data combination and cleaning:

1.  Combine all four raw data files into a single tibble, adding a categorical variable to track men's vs women's tour.

```{r}
#| message: false
#| warning: false
# Read in the four .csv files
M_Matches <- read_csv("RawData/charting-m-matches.csv")
M_Overview <- read_csv("RawData/charting-m-stats-Overview.csv")
W_Matches <- read_csv("RawData/charting-w-matches.csv")
W_Overview <- read_csv("RawData/charting-w-stats-Overview.csv")
# some error messages occurred for M_Matches and W_Matches during import
```

```{r}
# problems(M_Matches) indicates rows 907 and 908 of have issues
# print rows 906 to 909 to see if we can diagnose
M_Matches[906:909, ]
```
```{r}
# Row 907 has data issues, but also is redundant with row 908 
# Delete Row 907
M_Matches <- M_Matches[-907, ]
```

```{r}
# problems(W_Matches) indicates rows 1586 to 1594 have issues
# print rows 1585 to 1595 to see if we can diagnose
W_Matches[1585:1595, ]
```

```{r}
# Rows 1585 to 1594 have issues.  They have the wrong data in the wrong columns and/or are missing the surface data.
# Let's delete these rows.
W_Matches <- W_Matches[-(1585:1594), ]
```

```{r}
# Also variables Date and 'Best of' imported as chr while M_Matches has them as numeric.  Let's fix that.
W_Matches <- W_Matches |>
  mutate(Date = as.numeric(Date),
         `Best of` = as.numeric(`Best of`))
```


- Join each match table to the corresponding overview table

```{r}
M_Data <- left_join(M_Overview, M_Matches, by = "match_id")
W_Data <- left_join(W_Overview, W_Matches, by = "match_id")
```

- Combine the men's and women's data

```{r}
CombinedData <- bind_rows(M = M_Data, W = W_Data, .id = "Tour") # new "Tour" variable will be "M" or "W" depending on data source
```

2.  Delete variables and/or observations that will not be used in subsequent analysis

## 3. EDA Summary Tables

## 4. EDA Plots
