---
title: "ST 558 Project 2, Ryan Friedman"
format: html
editor: visual
---

This document will include a description of the dataset, detailed data combining/cleaning activities, and exploratory data analysis with summary tables and plots. These steps are prerequisite to creating the associated Shiny dashboard app.

```{r}
#| message: false
#| warning: false
library(janitor) # for clean_names()
library(tidyverse)
library(knitr)
library(gt)
library(fmsb) # for radarchart()
source("helpers.R")
```

## 1. Description of Dataset

The dataset for this analysis comes from the Tennis Match Charting Project (MCP), available at <https://github.com/JeffSackmann/tennis_MatchChartingProject.git>. The MCP aims to build a detailed public record of professional tennis matches. MCP data covers over 11,000 singles matches including both match-summary statistics and shot-by-shot detail. The project is open to the public both for accessing the data and for contributing new charted matches. This analysis uses only the summary data.  Note:  This is not a comprehensive dataset of all professional tennis matches. Because charting is done by individual volunteer contributors, the data tends to over-represent the biggest tournaments and the most popular players.

Four raw data .csv files are downloaded from the GitHub repo and placed in `RawData/`:

- `charting-m-matches.csv`

- `charting-m-stats-Overview.csv`

- `charting-w-matches.csv`

- `charting-w-stats-Overview.csv`

"m" and "w" indicate men's and women's professional tennis tour matches, respectively. "...-matches.csv" files contain general information on the matches played such as players, date, tournament, and surface. "...-stats-Overview.csv" files contain more detailed match performance information such as points won, aces, unforced errors, etc. Each observation across all four files contains a match_id variable, indicating the specific match observed. This variable will be used to join the data.

## 2. Data Combination and Cleaning

Combining and cleaning the data will involve the following general steps:

1.  Combine all four raw data files into a single tibble, adding a categorical variable to indicate 'M' or 'W' tour.

2.  Delete variables and/or observations that will not be used in subsequent analysis

3.  Add new variables calculated from existing variables

4.  Confirm correct format/type for all variables

The variables for the "-matches.csv" files are as follows:

| Variable | Type | Notes |
|------------------|------------------|------------------------------------|
| match_id | Identifier | Will use for join with overview files |
| Player 1 | Categorical | Name of Player 1 |
| Player 2 | Categorical | Name of Player 2 |
| Pl 1 hand | Categorical | Handed for Player 1 |
| Pl 2 hand | Categorical | Handed for Player 2 |
| Date | Numerical | Currently stored as YYYYMMDD integer. |
| Tournament | Categorical | Name of Tournament. Delete later. |
| Round | Categorical | Round of the Tournament (i.e. F, SF, R64, etc.). |
| Time | Categorical | Many missing values and inconsistent formatting. Delete later. |
| Court | Categorical | Name/number of court. Many missing values. Delete later. |
| Surface | Categorical | Hard, Clay or Grass |
| Umpire | Categorical | Name of umpire. Many missing values. Delete later. |
| Best of | Numerical | Best of 3 or Best of 5. Delete later.|
| Final TB? | Categorical | Meaning unclear. Delete later. |
| Charted by | Categorical | Name of data contributer. Delete later. |

The variables for the "-Overview.csv" files are as follows:

| Variable | Type | Notes |
|------------------|------------------|-------------------------------------|
| match_id | Identifier | Will use for join with overview files |
| player | Categorical | Name of player |
| set | Categorical | Stats by set number. Filter later to keep only "Total" and then delete. |
| serve_pts | Numerical | Points served |
| aces | Numerical | Aces |
| dfs | Numerical | Double faults |
| first_in | Numerical | First serves in play |
| first_won | Numerical | First serve points won |
| second_in | Numerical | Second serves in play |
| second_won | Numerical | Second serve points won |
| bk_pts | Numerical | Break points faced |
| bp_saved | Numerical | Break points saved |
| return_pts | Numerical | Points returned |
| return_pts_won | Numerical | Return points won |
| winners | Numerical | Winners |
| winners_fh | Numerical | Forehand winners |
| winners_bh | Numerical | Backhand winners |
| unforced | Numerical | Unforced errors |
| unforced_fh | Numerical | Forehand unforced errors |
| unforced_bh | Numerical | Backhand unforced errors |

Now to begin data combination and cleaning:

1.  Combine all four raw data files into a single tibble, adding a categorical variable to track men's vs women's tour.

```{r}
#| message: false
#| warning: false
# Read in the four .csv files
M_Matches <- read_csv("RawData/charting-m-matches.csv")
M_Overview <- read_csv("RawData/charting-m-stats-Overview.csv")
W_Matches <- read_csv("RawData/charting-w-matches.csv")
W_Overview <- read_csv("RawData/charting-w-stats-Overview.csv")
# some error messages occurred for M_Matches and W_Matches during import
```

```{r}
# problems(M_Matches) indicates rows 907 and 908 of have issues
# print rows 906 to 909 to see if we can diagnose
M_Matches[906:909, ]
```

```{r}
# Row 907 has data issues, but also is redundant with row 908 
# Delete Row 907
M_Matches <- M_Matches[-907, ]
```

```{r}
# problems(W_Matches) indicates rows 1586 to 1594 have issues
# print rows 1585 to 1595 to see if we can diagnose
W_Matches[1585:1595, ]
```

```{r}
# Rows 1585 to 1594 have issues.  They have the wrong data in the wrong columns and/or are missing the surface data.
# Let's delete these rows.
W_Matches <- W_Matches[-(1585:1594), ]
```

```{r}
# Also variables Date and 'Best of' imported as chr while M_Matches has them as numeric.  Let's fix that.
W_Matches <- W_Matches |>
  mutate(Date = as.numeric(Date),
         `Best of` = as.numeric(`Best of`))
```

- Join each match table to the corresponding overview table

```{r}
M_Data <- left_join(M_Overview, M_Matches, by = "match_id")
W_Data <- left_join(W_Overview, W_Matches, by = "match_id")
```

- Combine the men's and women's data

```{r}
CombinedData <- bind_rows(M = M_Data, W = W_Data, .id = "Tour") # new "Tour" variable will be "M" or "W" depending on data source
```

2.  Delete variables and/or observations that will not be used in subsequent analysis

- First delete rows and columns as described earlier. Also fix remaining names

```{r}
CombinedData <- CombinedData |>
  filter(set == "Total") |> # only include full match observations (exclude per set observations)
  select(-Time, # deleting variables as described earlier
         -Court,
         -Umpire,
         -`Final TB?`,
         -`Charted by`,
         -set) |>
  clean_names() # to correct variable names with spaces and non-standard characters
```

- Each row represents a single player for a particular match. As such, "player_1" and "player_2" are redundant with "player". However they do correlate with "pl_1_hand" and "pl_2_hand". Let's delete those four variables while creating new variable "handed" that indicates whether "player" is left or right-handed.

```{r}
CombinedData <- CombinedData |>
  mutate(handed = case_when(
    player == player_1 ~ pl_1_hand, # if player is player_1, assign pl_1_hand
    player == player_2 ~ pl_2_hand, # if player is player_2, assign pl_2_hand
    TRUE ~ NA_character_ # otherwise (i.e. name mismatch or missing assign NA)
  )) |>
  select (-player_1,
          -player_2,
          -pl_1_hand,
          -pl_2_hand) # deleting variables no longer needed
```

3.  Add new variables calculated from existing variables

- Percentage variables are more useful for analysis than count variables, because they are less dependent upon the length of the match.

```{r}
CombinedData <- CombinedData |>
  mutate(total_pts = serve_pts + return_pts, # total points played
         first_in_pct = first_in / serve_pts, # first serve in percentage
         ace_pct = aces / serve_pts, # percentage of service points that are aces
         first_won_pct = first_won / first_in, # percentage of points won on first serve
         second_won_pct = second_won / (second_in + dfs), # percentage of points won on second serve
         total_won_pct = (first_won + second_won + return_pts_won) / total_pts, # percentage of total points won
         df_pct = dfs / (serve_pts - first_in), # double fault percentage
         bp_saved_pct = bp_saved / bk_pts, # break point save percentage
         serve_pts_won_pct = (first_won + second_won) / serve_pts, # percentage of service points won
         return_pts_won_pct = return_pts_won / return_pts, # percentage of return points won
         winner_pct = winners / total_pts, # percentage of points ending in winner
         unforced_pct = unforced / total_pts, # percentage of points ending in unforced error
         winner_fh_pct = winners_fh / total_pts, # percentage of points ending in forehand winner
         winner_bh_pct = winners_bh / total_pts, # percentage of points ending in backhand winner
         unforced_fh_pct = unforced_fh / total_pts, # percentage of points ending in forehand unforced error
         unforced_bh_pct = unforced_bh / total_pts, # percentage of points ending in backhand unforced error
         fh_eff_pct = winner_fh_pct - unforced_fh_pct, # a measure of forehand effectiveness
         bh_eff_pct = winner_bh_pct - unforced_bh_pct, # a measure of backhand effectiveness
         aggression_pct = winner_pct + unforced_pct) # a measure of overall aggression
```

4.  Confirm correct format/type for all variables

- Lets look for any remaining possible data issues. Use summary() to evaluate all variables.
```{r}
summary(CombinedData)
```

- "surface" is showing 4 unique levels.  3 is expected.  Find and correct the error.

```{r}
unique(CombinedData$surface) # check unique elements of surface
```

```{r}
# Filer to exclude that particular row.  Keep NA rows. 
CombinedData <- CombinedData |> filter(surface != "Eva Asderaki-Moore" | is.na(surface))
```

- "handed" is showing 3 unique levels.  2 is expected.  Find and correct the error.

```{r}
unique(CombinedData$handed) # check unique elements of handed
```
```{r}
# Correct "l" to "L"
CombinedData <- CombinedData |> mutate(handed = toupper(handed))
```

- "round" has some unexpected and uncommon values.  Review and trim appropriately

```{r}
print(rows=14, CombinedData |> 
  drop_na(round) |>
  group_by(round) |>
  summarize(count = n())) # get counts of all 14 levels of round
```

```{r}
# Drop PO and PQ as they are very uncommon.  Keep NA
CombinedData <- CombinedData |>
  filter(!round %in% c("BR", "PO", "PQ") | is.na(round))
```

- NA values for the "pct_" variables are expected, as the denominator for the calculation may at times be zero.

- "tournament" and "best_of" are no longer desired.  Delete these.

```{r}
CombinedData <- CombinedData |>
  select(-tournament,
         -best_of)
```

- Convert "date" variable to "year".  For the purpose of this analysis, there is no need for any detail beyond the year.

```{r}
CombinedData <- CombinedData |>
  mutate(year = as.numeric(substr(as.character(date), 1, 4))) |> # convert "date" to a string, keep only the first four characters, then convert back to number
  select(-date) # remove "date"
```

- Convert select categorical variables to factors:  "tour" "surface" "handed" and "round"

```{r}
CombinedData <- CombinedData |>
  mutate(tour = factor(tour),
         surface = factor(surface),
         handed = factor(handed),
         round = factor(round,
                        levels = c("Q1","Q2","Q3","R128","R64","R32","R16","QF","RR","SF","F"),
                        ordered = TRUE), # putting the levels of round in logical order based on proximity to winning the tournament
         round_group = factor(round_group[as.character(round)],
                              levels = round_group_levels,
                              ordered = TRUE))
```

- Delete large items from the environment that are no longer needed

```{r}
rm(M_Data,
   M_Matches,
   M_Overview,
   W_Data,
   W_Matches,
   W_Overview)
```

- Now save the cleaned data as an .rds file that can be read by the Shiny app later

```{r}
saveRDS(CombinedData, "CombinedData.rds")
```

## 3. EDA Summary Tables

- Create a one-way contingency table for "handed" "tour" "surface" and "round" variables

```{r}
kable(CombinedData |> 
  drop_na(handed) |>
  group_by(handed) |>
  summarize(count = n()))
# Right-handed players are more commonly observed in the dataset than left handed players by about 8.5:1 

kable(CombinedData |> 
  drop_na(tour) |>
  group_by(tour) |>
  summarize(count = n()))
# The dataset includes nearly twice as many matches from the Men's tour than the Women's tour

kable(CombinedData |> 
  drop_na(surface) |>
  group_by(surface) |>
  summarize(count = n()))
# Clay courts represent about twice as many matches in the dataset as Grass courts.
# Hard courts represent nearly 3 times as many matches in the dataset as Clay courts.

kable(CombinedData |> 
  drop_na(round) |>
  group_by(round) |>
  summarize(count = n()))
# In a standard single-elimination tournament, there is 1 F match, 2 SF matches, 4 QF matches, etc.
# However, this dataset includes roughly equal counts (~3,000 each) of F, SF, and QF 
# This confirms the suspected sampling bias towards higher-profile matches
```

- Create a two-way contingency table for surface vs tour and handed vs tour

```{r}
#| message: false
kable(CombinedData |>
        drop_na(handed, tour) |>
        group_by(handed, tour) |>
        summarize(count = n()) |>
        pivot_wider(names_from = tour, values_from = count))
# Left-handed players are the minority on both tours, with men's tennis showing a slightly higher share

kable(CombinedData |> 
        drop_na(surface, tour) |>
        group_by(surface, tour) |>
        summarize(count = n()) |>
        pivot_wider(names_from = tour, values_from = count))
# Hard courts are most popular, and there do not appear to be any notable differences in surface proportions between the two tours
```

- Find measures of center and spread for ace_pct and serve_pts_won_pct by tour

```{r}
CombinedData |>
  drop_na(tour) |>
  group_by(tour) |>
  summarize(across(.cols = c(ace_pct, serve_pts_won_pct),
                   list("mean" = ~ mean(.x, na.rm = TRUE),
                        "median" = ~ median(.x, na.rm = TRUE),
                        "sd" = ~ sd(.x, na.rm = TRUE),
                        "Q1" = ~ quantile(.x, 0.25, na.rm = TRUE),
                        "Q3" = ~ quantile(.x, 0.75, na.rm = TRUE)),
                   .names = "{.fn}.{.col}")) |> # creates names from the variable and the statistic
  pivot_longer(
    cols = -tour, # exclude tour from the pivot
    names_to = c("Statistic", "Variable"),
    names_sep = "\\.") |>
  pivot_wider(
    names_from = Statistic,
    values_from = value) |>
  gt(groupname_col = "tour", rowname_col = "Variable") |> # organizes the Variable column by tour and removes the "Variable" title
  fmt_number(decimals = 3) |> # rounds everything to 3 decimals
  tab_header(
    title = "Ace Percentage and Serve Points Won Percentage: Mean, Median, SD, Q1, and Q3",
    subtitle = "Stratified by tour")
# This table illustrates the greater weight of serve performance on the Men's tour vs the Women's tour.
# Mean ace percentage is roughly double on the Men's tour (8.4% vs 4.2%).
# On the Men's tour, even the bottom quartile wins most serve points (Q1 = 58.2%), a clear serve advantage.
# On the Women's tour, the bottom quartile is near break-even (Q1 = 50.7%), so the serve advantage nearly vanishes for lower-serving performances.
```

- Find measures of center and spread for winner_pct and unforced_pct by round

```{r}
CombinedData |>
  drop_na(round) |>
  group_by(round) |>
  summarize(across(.cols = c(winner_pct, unforced_pct),
                   list("mean" = ~ mean(.x, na.rm = TRUE),
                        "median" = ~ median(.x, na.rm = TRUE),
                        "sd" = ~ sd(.x, na.rm = TRUE),
                        "Q1" = ~ quantile(.x, 0.25, na.rm = TRUE),
                        "Q3" = ~ quantile(.x, 0.75, na.rm = TRUE)),
                   .names = "{.fn}.{.col}")) |> # creates names from the variable and the statistic
  pivot_longer(
    cols = -round, # exclude round from the pivot
    names_to = c("Statistic", "Variable"),
    names_sep = "\\.") |>
  pivot_wider(
    names_from = Statistic,
    values_from = value) |>
  gt(groupname_col = "round", rowname_col = "Variable") |> # organizes the Variable column by round and removes the "Variable" title
  fmt_number(decimals = 3) |> # rounds everything to 3 decimals
  tab_header(
    title = "Winner Percentage and Unforced Error Percentage: Mean, Median, SD, Q1, and Q3",
    subtitle = "Stratified by round")
# This table examines whether play "quality" shifts from qualifying rounds through the final.
# Winner percentage rises gradually from qualifying (~14%) to a plateau around the QF-SF (~16.7%).
# Unforced error percentage drifts slightly downward, from ~19% in qualifying to ~17.4-17.9% in later rounds.
```

## 4. EDA Plots

- Create a bar chart for surface

```{r}
g <- ggplot(data = CombinedData |> drop_na(surface), aes(x = surface, fill = surface))
g + geom_bar() +
    labs(x = var_labels[["surface"]], # reference corresponding label
         title = paste("Count of Charted Matches by", var_labels[["surface"]])) + 
  theme(plot.title = element_text(hjust = 0.5), # centering the title
        legend.position = "none") # remove redundant legend
# The majority of matches in the dataset are played on hard court, followed by clay and then grass.
```

- Create a side-by-side bar chart for tour and handed

```{r}
g <- ggplot(data = CombinedData |> drop_na(tour, handed), aes(x = tour, fill = handed))
g + geom_bar(position = "dodge") +
    labs(x = var_labels[["tour"]], 
         title = paste("Count of Charted Matches by", var_labels[["tour"]], "and", var_labels[["handed"]])) + 
  scale_fill_discrete(var_labels[["handed"]]) +
  theme(plot.title = element_text(hjust = 0.5))
# Right-handed players outnumber left-handed players, with no notable difference in representation between Men's and Women's tours
```

- Create a histogram for year, faceted by tour

```{r}
g <- ggplot(CombinedData |> drop_na(year, tour), aes(x = year))
g + geom_histogram(binwidth = 5, alpha = .5, aes(fill = tour)) + 
  facet_wrap(~ tour) +
  labs(x = var_labels[["year"]],
       fill = var_labels[["tour"]],
       title = paste("Charted Match", var_labels[["year"]]),
       subtitle = paste("Faceted by", var_labels[["tour"]])) + 
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5), # Centers the title and subtitle
        legend.position = "none") # remove redundant legend
# The majority of matches in the dataset are from the early 2010's and later
# Overall, there are more men's matches than women's matches by about 2:1
# However prior to the 2010's, men's matches outnumber women's matches by more than
# Note: this data is only representative of the specific matches charted in the dataset (i.e. not a comprehensive representation of the history of professional tennis)
```

- Create a side-by-side box plot for return_pts_won_pct by surface

```{r}
g <- ggplot(CombinedData |> drop_na(return_pts_won_pct, surface))
g + geom_boxplot(aes(x = surface, y = return_pts_won_pct, fill = surface)) +
  labs(x = var_labels[["surface"]], # reference corresponding label for surface
       y = var_labels[["return_pts_won_pct"]],
       title = paste(var_labels[["return_pts_won_pct"]], "by", var_labels[["surface"]])) + 
  theme(plot.title = element_text(hjust = 0.5), # Centers the title
        legend.position = "none") # remove redundant legend
# The data indicates that Winning return points is more difficult on grass and less difficult on clay.
# This aligns with expectations.  Clay courts have high and slower bounces while grass courts have low and skidding bounces, affecting return difficulty.
```

- Create a scatter plot for winners_pct vs unforced_pct by tour

```{r}
g <- ggplot(CombinedData |>
              drop_na(winner_pct, unforced_pct, tour),
            aes(x = winner_pct, y = unforced_pct, color = tour))
g + geom_point(alpha = 0.5, size = 0.9) + 
  labs(x = var_labels[["winner_pct"]],
       y = var_labels[["unforced_pct"]],
       title = paste(var_labels[["unforced_pct"]], "vs", var_labels[["winner_pct"]], "by", var_labels[["tour"]]),
       color = var_labels[["tour"]]) + 
  theme(plot.title = element_text(hjust = 0.5)) # Centers the title
# The men's tour tends to have a slightly higher percentage of points ending in winners than the women's tour
# This is likely in large part due to an increased frequency of aces (which are counted as winners in tennis statistics)
# There is no obvious different in percentage of points ending in unforced errors.
```

- Create a radar chart for Rafael Nadal vs John Isner comparing means for the following variables: total_won_pct, ace_pct, serve_pts_won_pct, return_pts_won_pct, aggression_pct, fh_eff_pct, and bf_eff_pct

```{r}

# first let's create a new table containing the means of the variables of interest, grouped by player
player_means <- CombinedData |>
  group_by(player) |>
  summarize(across(c(total_won_pct, ace_pct, serve_pts_won_pct, return_pts_won_pct,
                     aggression_pct, fh_eff_pct, bh_eff_pct),
                   ~ mean(.x, na.rm = TRUE)), # calculates the mean of each variable by player
            n = n()) # adds n, the qty of matches per player 

# radarchart() will need to know the min and max of each variable for scaling purposes, so let's take a look
summary(player_means)
```
```{r}
# there are some outlier values here, most likely due to low n for those players
# let's revisit while filtering out players with less than 10 matches observed
player_means |> filter(n >= 10) |> summary()
```
```{r}
# the above provides a much more reasonable picture
# now we can manually construct a radar_bounds dataframe, rounding where desired
# this code will also exist in helpers.R for use later

radar_bounds <- data.frame(
  total_won_pct      = c(0.60, 0.40),
  ace_pct            = c(0.25, 0.00),
  serve_pts_won_pct  = c(0.75, 0.45),
  return_pts_won_pct = c(0.55, 0.20),
  aggression_pct     = c(0.50, 0.20),
  fh_eff_pct         = c(0.05, -0.08),
  bh_eff_pct         = c(0.03, -0.10)
)
rownames(radar_bounds) <- c("max", "min")
```

```{r}
# now make a function to build the df that feeds radarchart()
# this will also live in helpers.R
build_radar_data <- function(data, player_a, player_b) {
  player_rows <- data |> # getting the player data
    filter(player %in% c(player_a, player_b)) |> # select the two players
    group_by(player) |>
    summarize(across(c(total_won_pct, ace_pct, serve_pts_won_pct, return_pts_won_pct,
                       aggression_pct, fh_eff_pct, bh_eff_pct),
                     ~ mean(.x, na.rm = TRUE)),# calculating the means of each variable for each player
              ) |> 
    as.data.frame() # radarchart() needs data.frame format
  
  rownames(player_rows) <- player_rows$player # add the players as row names, which is how radarchart() will get these labels
  player_rows$player <- NULL # column of player names no longer needed
  
  
  out <- rbind(radar_bounds, player_rows) # combine the variable min/max with the player data into one data frame
  colnames(out) <- radar_labels[colnames(out)] # uses the proper variable labels from radar_labels in helpers.R
  out # return out
}

# run function for the two players of interest
radardata<-build_radar_data(CombinedData, "Rafael Nadal", "John Isner")
radardata
```

```{r}
par(mar = c(1, 1, 1, 1)) # narrow margins for less white space
radarchart(radardata,
           vlcex = 0.8, # scales label text size
           pcol = c("red", "blue"), # applies colors
           pfcol = c(adjustcolor("red", alpha.f = 0.3), # applies fill colors with transparency
                     adjustcolor("blue", alpha.f = 0.3)), 
           plwd = 2, # line width
           plty = 1, # solid lines
           cglcol = "grey", # grid line color
           cglty = 1, # solid grid line
           cglwd = 0.5, # grid line width
           title = "Mean Metrics Across All Charted Matches")
legend("topright", # build legend in top right
       legend = rownames(radardata)[-c(1, 2)],   # drop max/min rows
       col = c("red", "blue"), # legend colors
       lwd = 4, # legend line width
       bty = "n", # remove legend border box
       cex = 0.8) # scale legend text size
```
```{r}
# In the dashboard, it would be good to have a table of this data below the radarchart
# Create a function to build this radar chart companion table 
# this will also live in helpers.R
build_radar_table <- function(data, player_a, player_b) {
  data |>
    filter(player %in% c(player_a, player_b)) |> # filters for the two players
    group_by(player) |>
    summarize(n = n(),
              across(c(total_won_pct, ace_pct, serve_pts_won_pct, return_pts_won_pct,
                       aggression_pct, fh_eff_pct, bh_eff_pct),
                     ~ mean(.x, na.rm = TRUE)),
              .groups = "drop") |>
    rename_with(~ var_labels[.x], .cols = -c(player, n)) |> # use var_labels for renaming all variables except n and player
    rename(Player = player, `Charted Matches` = n) # manually renaming n and player
}

build_radar_table(CombinedData, "Rafael Nadal", "John Isner") |>
  gt() |>
  fmt_number(columns = -c(Player, `Charted Matches`), decimals = 3) |>
  tab_header(title = "Mean Player Metrics",
             subtitle = "Across All Charted Matches") 
```

