---
title: "Project 2"
format: html
editor: visual
---

This markdown document will include a description of the dataset, detailed data combining/cleaning activities, and exploratory data analysis with summary tables and plots. These steps are prerequisite to creating the associated Shiny dashboard app.

```{r}
#| message: false
#| warning: false
library(janitor)
library(tidyverse)
```

## 1. Description of Dataset

The dataset for this analysis comes from the Tennis Match Charting Project (MCP), available at <https://github.com/JeffSackmann/tennis_MatchChartingProject.git>. The MCP aims to build a detailed public record of professional tennis matches. MCP data covers over 11,000 singles matches including both match-summary statistics and shot-by-shot detail. The project is open to the public both for accessing the data and for contributing new charted matches. This analysis uses only the summary data.

Four raw data .csv files are downloaded from the GitHub repo and placed in `RawData/`:

- `charting-m-matches.csv`

- `charting-m-stats-Overview.csv`

- `charting-w-matches.csv`

- `charting-w-stats-Overview.csv`

"m" and "w" indicate men's and women's professional tennis tour matches, respectively. "...-matches.csv" files contain general information on the matches played such as players, date, tournament, and surface. "...-stats-Overview.csv" files contain more detailed match performance information such as points won, aces, unforced errors, etc. Each observation across all four files contains a match_id variable, indicating the specific match observed. This variable will be used to join the data.

## 2. Data Combination and Cleaning

Combining and cleaning the data will involve the following general steps:

1.  Combine all four raw data files into a single tibble, adding a categorical variable to indicate 'M' or 'W' tour.

2.  Delete variables and/or observations that will not be used in subsequent analysis

3.  Add new variables calculated from existing variables

4.  Confirm correct format/type for all variables

The variables for the "-matches.csv" files are as follows:

| Variable | Type | Notes |
|------------------|------------------|------------------------------------|
| match_id | Identifier | Will use for join with overview files |
| Player 1 | Categorical | Name of Player 1 |
| Player 2 | Categorical | Name of Player 2 |
| Pl 1 hand | Categorical | Handed for Player 1 |
| Pl 2 hand | Categorical | Handed for Player 2 |
| Date | Numerical | Currently stored as YYYYMMDD integer. |
| Tournament | Categorical | Name of Tournament. Delete later. |
| Round | Categorical | Round of the Tournament (i.e. F, SF, R64, etc.).  Delete later |
| Time | Categorical | Many missing values and inconsistent formatting. Delete later. |
| Court | Categorical | Name/number of court. Many missing values. Delete later. |
| Surface | Categorical | Hard, Clay or Grass |
| Umpire | Categorical | Name of umpire. Many missing values. Delete later. |
| Best of | Numerical | Best of 3 or Best of 5. Delete later.|
| Final TB? | Categorical | Meaning unclear. Delete later. |
| Charted by | Categorical | Name of data contributer. Delete later. |

The variables for the "-Overview.csv" files are as follows:

| Variable | Type | Notes |
|------------------|------------------|-------------------------------------|
| match_id | Identifier | Will use for join with overview files |
| player | Categorical | Name of player |
| set | Categorical | Stats by set number. Filter later to keep only "Total" and then delete. |
| serve_pts | Numerical | Points served |
| aces | Numerical | Aces |
| dfs | Numerical | Double faults |
| first_in | Numerical | First serves in play |
| first_won | Numerical | First serve points won |
| second_in | Numerical | Second serves in play |
| second_won | Numerical | Second serve points won |
| bk_pts | Numerical | Break points faced |
| bp_saved | Numerical | Break points saved |
| return_pts | Numerical | Points returned |
| return_pts_won | Numerical | Return points won |
| winners | Numerical | Winners |
| winners_fh | Numerical | Forehand winners |
| winners_bh | Numerical | Backhand winners |
| unforced | Numerical | Unforced errors |
| unforced_fh | Numerical | Forehand unforced errors |
| unforced_bh | Numerical | Backhand unforced errors |

Now to begin data combination and cleaning:

1.  Combine all four raw data files into a single tibble, adding a categorical variable to track men's vs women's tour.

```{r}
#| message: false
#| warning: false
# Read in the four .csv files
M_Matches <- read_csv("RawData/charting-m-matches.csv")
M_Overview <- read_csv("RawData/charting-m-stats-Overview.csv")
W_Matches <- read_csv("RawData/charting-w-matches.csv")
W_Overview <- read_csv("RawData/charting-w-stats-Overview.csv")
# some error messages occurred for M_Matches and W_Matches during import
```

```{r}
# problems(M_Matches) indicates rows 907 and 908 of have issues
# print rows 906 to 909 to see if we can diagnose
M_Matches[906:909, ]
```

```{r}
# Row 907 has data issues, but also is redundant with row 908 
# Delete Row 907
M_Matches <- M_Matches[-907, ]
```

```{r}
# problems(W_Matches) indicates rows 1586 to 1594 have issues
# print rows 1585 to 1595 to see if we can diagnose
W_Matches[1585:1595, ]
```

```{r}
# Rows 1585 to 1594 have issues.  They have the wrong data in the wrong columns and/or are missing the surface data.
# Let's delete these rows.
W_Matches <- W_Matches[-(1585:1594), ]
```

```{r}
# Also variables Date and 'Best of' imported as chr while M_Matches has them as numeric.  Let's fix that.
W_Matches <- W_Matches |>
  mutate(Date = as.numeric(Date),
         `Best of` = as.numeric(`Best of`))
```

- Join each match table to the corresponding overview table

```{r}
M_Data <- left_join(M_Overview, M_Matches, by = "match_id")
W_Data <- left_join(W_Overview, W_Matches, by = "match_id")
```

- Combine the men's and women's data

```{r}
CombinedData <- bind_rows(M = M_Data, W = W_Data, .id = "Tour") # new "Tour" variable will be "M" or "W" depending on data source
```

2.  Delete variables and/or observations that will not be used in subsequent analysis

- First delete rows and columns as described earlier. Also fix remaining names

```{r}
CombinedData <- CombinedData |>
  filter(set == "Total") |> # only include full match observations (exclude per set observations)
  select(-Time, # deleting variables as described earlier
         -Court,
         -Umpire,
         -`Final TB?`,
         -`Charted by`,
         -set) |>
  clean_names() # to correct variable names with spaces and non-standard characters
```

- Each row represents a single player for a particular match. As such, "player_1" and "player_2" are redundant with "player". However they do correlate with "pl_1_hand" and "pl_2_hand". Let's delete those four variables while creating new variable "handed" that indicates whether "player" is left or right-handed.

```{r}
CombinedData <- CombinedData |>
  mutate(handed = case_when(
    player == player_1 ~ pl_1_hand, # if player is player_1, assign pl_1_hand
    player == player_2 ~ pl_2_hand, # if player is player_2, assign pl_2_hand
    TRUE ~ NA_character_ # otherwise (i.e. name mismatch or missing assign NA)
  )) |>
  select (-player_1,
          -player_2,
          -pl_1_hand,
          -pl_2_hand) # deleting variables no longer needed
```

3.  Add new variables calculated from existing variables

- Percentage variables are more useful for analysis than count variables, because they are less dependent upon the length of the match.

```{r}
CombinedData <- CombinedData |>
  mutate(total_pts = serve_pts + return_pts, # total points played
         first_in_pct = first_in / serve_pts, # first serve in percentage
         ace_pct = aces / serve_pts, # percentage of service points that are aces
         first_won_pct = first_won / first_in, # percentage of points won on first serve
         second_won_pct = second_won / (second_in + dfs), # percentage of points won on second serve
         total_won_pct = (first_won + second_won + return_pts_won) / total_pts, # percentage of total points won
         df_pct = dfs / (serve_pts - first_in), # double fault percentage
         bp_saved_pct = bp_saved / bk_pts, # break point save percentage
         serve_pts_won_pct = (first_won + second_won) / serve_pts, # percentage of service points won
         return_pts_won_pct = return_pts_won / return_pts, # percentage of return points won
         winner_pct = winners / total_pts, # percentage of points ending in winner
         unforced_pct = unforced / total_pts, # percentage of points ending in unforced error
         winner_fh_pct = winners_fh / total_pts, # percentage of points ending in forehand winner
         winner_bh_pct = winners_bh / total_pts, # percentage of points ending in backhand winner
         unforced_fh_pct = unforced_fh / total_pts, # percentage of points ending in forehand unforced error
         unforced_bh_pct = unforced_bh / total_pts) # percentage of points ending in backhand unforced error
```

4.  Confirm correct format/type for all variables

- Lets look for any remaining possible data issues. Use summary() to evaluate all variables.
```{r}
summary(CombinedData)
```

- "surface" is showing 4 unique levels.  3 is expected.  Find and correct the error.

```{r}
unique(CombinedData$surface) # check unique elements of surface
```

```{r}
# Filer to exclude that particular row.  Keep NA rows. 
CombinedData <- CombinedData |> filter(surface != "Eva Asderaki-Moore" | is.na(surface))
```

- "handed" is showing 3 unique levels.  2 is expected.  Find and correct the error.

```{r}
unique(CombinedData$handed) # check unique elements of handed
```
```{r}
# Correct "l" to "L"
CombinedData <- CombinedData |> mutate(handed = toupper(handed))
```

- NA values for the "pct_" variables are expected, as the denominator for the calculation may at times be zero.

- "tournament" "round" and "best_of" are no longer desired.  Delete these.

```{r}
CombinedData <- CombinedData |>
  select(-tournament,
         -round,
         -best_of)
```

- Convert "date" variable to "year".  For the purpose of this analysis, there is no need for any detail beyond the year.

```{r}
CombinedData <- CombinedData |>
  mutate(year = as.numeric(substr(as.character(date), 1, 4))) |> # convert "date" to a string, keep only the first four characters, then convert back to number
  select(-date) # remove "date"
```

- Convert select categorical variables to factors:  "tour" "surface" "handed"

```{r}
CombinedData <- CombinedData |>
  mutate(tour = factor(tour),
         surface = factor(surface),
         handed = factor(handed))
```

- Delete large items from the environment that are no longer needed

```{r}
rm(M_Data,
   M_Matches,
   M_Overview,
   W_Data,
   W_Matches,
   W_Overview)
```

- Now save the cleaned data as an .rds file that can be read by the Shiny app later

```{r}
saveRDS(CombinedData, "CombinedData.rds")
```

## 3. EDA Summary Tables

## 4. EDA Plots
